---
title: A very short introduction to pipes 
weight: 2
---

<style>
body {
text-align: justify}
</style>
  
Here we introduce the pipe operator, which is provided by the [`magrittr`](https://cran.r-project.org/web/packages/magrittr/) `R` package. This introduction is extremely short, and we refer to [R for Data Science](https://r4ds.had.co.nz) for details.
  
## Pipe basic

The pipe operator `%>%` is quite simple, but very useful in certain situations. Consider the following code:
```{r, fig.show = TRUE}
x <- seq(0, 2*pi, by = 0.01)

plot(sqrt(abs(cos(x))))
```
which is plotting $\sqrt(|\text{cos}(x)|)$. The piped version of this is:
```{r}
library(magrittr)

x %>% cos %>% abs %>% sqrt %>% plot
```
As you can see, the output is the same, with the exception of the y-axis label, which is now "." (we'll explain the reason for that in a moment). Now, what is happening? Essentially, the pipe operator is used in expressions of the type `R_object %>% A_function` and it trasforms this expression into something equivalent to `A_function( R_object )`. So the function composition `f1(f2(x))` becomes `x %>% f2 %>% f1`.  

To be more precise, the code `x %>% cos %>% abs %>% sqrt %>% plot` is equivalent to:
```{r}
 myFun <- function(.){ 
  
 . <- cos( . )
 . <- abs( . )
 . <- sqrt( . )
 plot( . )
 
}
 
myFun(x)
```
that is, we are applying a sequence of functions and we are storing the partial results in the ".". This should explain why `.` appears in the y-axis label. 

Now, what do we do if we want to specify if we want to specify some extra arguments, for example as in the code
```{r}
plot(x = x, y = sqrt(abs(cos(x))), ylab = "y")
```
? This is achieved as follows:
```{r}
x %>% cos %>% abs %>% sqrt %>% plot(x = x, y = ., ylab = "y")
```
Hence, we use `.` as a placeholder for the argument that is being piped (the lhs of the pipe operator). 

WARNING: a clarifications is needed here. By default the placeholder `.` will be used as the first argument of the function to be applied (here `plot()`). That is, `x %>% plot(ylab = "y")` is equivalent to `plot(., ylab = "y")` where `. == x`. The behaviour is overridden if the placeholders appear elsewhere on the rhs. For instance `x %>% plot(x = ., ylab = "y")` is not equivalent to `plot(x, x = ., ylab = "y")`, but it is the same as `plot(x = x, ylab = "y")`. However the following does not work:
```{r, eval = FALSE}
x %>% plot(x = x, y = . + 1, ylab = "y")
# Error in plot.xy(xy, type, ...) : invalid plot type
```
Why? Because here the placeholder appears only the nested expression (`. + 1`) and the above code is equivalent to `plot(x, x = x, y = x + 1, ylab = "y")`, which leads to an error. Hence, when the placeholder appears only in nested expressions on the rhs of the pipe, the `%>%` follows the default behaviour consisting using the `.` (where `. == x` here) as the first argument of the function to be applied (`plot`). To override this we must sandwhich the rhs of the pipe using curly `{}` brackets:
```{r}
x %>% { plot(x = x, y = . + 1, ylab = "y") }
```
which makes so that the `.` appears only where in it is explicitly used.
 
Having clarified this, we can now show how the `.` can be used multiple times inside the rhs function, for instance:
```{r}
3 %>% { matrix(1 : (. * .), ncol = ., nrow = .) }
```
is equivalent to `matrix(1 : (3 * 3), ncol = 3, nrow = 3)`. Also, we can split up the lhs of the pipe into its elements, for instance:
```{r}
x <- list(letters[1:6], 2, 3, TRUE)

x %>% { matrix(.[[1]], nrow = .[[2]], ncol = .[[3]], byrow = .[[4]]) }
```
which is equivalent to `matrix(letters[1:6], nrow = 2, ncol = 3, byrow = TRUE)`.





<!-- Consider the data set: -->
<!-- ```{r} -->
<!-- library(qgam) -->
<!-- data(UKload) -->

<!-- # plot(transform(head(subset(UKload, Dow == "lundi", select = c("NetDemand", )), 100), Posan == Posan * 365) -->
<!-- ``` -->
<!-- A value can be assigned to a variable using the assignment operator `<-`. The variable is created if it doesn't already exist. -->

