---
title: "3. A ggplot2 case study: the mgcViz package"
weight: 2
---

<style>
body {
text-align: justify}
</style>

As we said before, `ggplot2` provides a powerful framework for building visualization libraries. The best way to see why this is the case, might to examine a real example. Here we consider the `mgcViz` package, with provides a layered framework for visualizing Generalized Additive Models (GAMs) using `ggplot2`. You might have heard about GAMs before but, if you didn't, don't worry because a good background on GAMs is not necessary to understand this section. 

GAMs are an extension of Generalized Linear Models (GLMs), which allow using spline-based smooth effects in addition to the usual linear effects used in GLMs. For example, a Poisson GLM might have the following structure $x$:
$$
y \sim \text{Pois}\{\mu({\bf x})\}, \;\;\; \log \mu({\bf x}) = \beta_0 + \beta_1 x_1 + \beta_2 x_2,
$$
where ${\bf x} = \{x_1, x_2\}$ and $\beta_0, \beta_1, \beta_2$ are regression coefficients. So we are modelling the log-rate of a Poisson distribution, using linear effects for variables $x_1$ and $x_2$. A GAM extension of such model could be: 
$$
\log \mu({\bf x}) = \beta_0 + f_1(x_1) + f_2(x_2),
$$
where $f_1$ and $f_2$ are non-linear functions, built using spline basis expansions. That is the first effect is:
$$
f_1(x_1) = \sum_{j=1}^{K_1} \beta^1_j b_j^1(x_1), 
$$
where $b_1^1(x_1), \dots, b_{K_1}^1(x_1)$ are known basis functions and $\beta_1^1, \dots, \beta_{K_1}^1$ are unknown regression coefficients (to be estimated). Effect $f_2$ is defined similarly. There are many classes of effects, for example they can be multidimensional, such as:
$$
\log \mu({\bf x}) = \beta_0 + f(x_1, x_2),
$$
where $f(x_1, x_2)$ is built using 2-D basis functions.

Here we are not interested in discussing how GAMs are estimated, what they can be used for, etc... but we want to demonstrate the advantages of adopting a layer-based system for GAM models. Consider again the following data sets:
```{r, message = FALSE}
library(qgam)
data(UKload)
head(UKload)
```

See `?UKload` for more information about the data. A simple GAM model for aggregate electricity demand $\text{Dem_t}$ could be:
$$
\mathbb{E}(\text{Dem}_t) \sim \psi_{D_t} + f_1(T_t) + f_2(T^s_t) + f_3(S_t) + f_4(L_{t-48}) + f_5(t),
$$
where we have smooth effects for the hourly temperatures ($T_t$), the smoothed temperature ($T^s_t$, defined by $T_{t}^s = \alpha T_{t-1}^s + (1-\alpha)T_t$ with $\alpha = 0.95$), the variable indicating the position within the year ($S_t$), the sequential index representing time ($t$) and the observed load at the same time of the previous day ($L_{t-48}$). $\psi_{D_t}$ is a parametric effect, whose value depends on $D_t$, which is the factor variable indicating the day of the week. This model can be fitted using `mgcv` as follows:
```{r}
fitG <- gam(NetDemand ~ Dow + s(wM) + s(wM_s95) + s(Posan) + 
                        s(NetDemand.48) + s(Trend, k = 6), data = UKload)
```
Let's say that we now want to visualize the seasonal effect, that is the effect of $S_t$ (`Posan` in the code). This is done as follows:
```{r, message = F, fig.width = 6, fig.height = 4, fig.align='center'}
plot(fitG, select = 3, scale = FALSE)
```
where the `select` argument is used to determine which smooth effect should be plotted. Given that `plot` is a generic function and that
```{r}
class(fitG)
```
the above code is calling `plot.gam` from the `mgcv` package. `plot.gam` is based on base R plotting functions, not on `ggplot2`. This function does its job, but has some limitations that can be addressed by adopting a layer-based plotting framework. The first issue is that `plot.gam` has quite a lot of arguments:
```{r}
args(plot.gam)
```
Furthermore, many of its arguments are not used at all when most function calls, for instance
```{r, message = F, fig.width = 6, fig.height = 4, fig.show='hide'}
plot(fitG, select = 3, scale = FALSE, n2 = 100, theta = 10, n3 = 4)
```

produces exactly the same plot as before (not shown), because these arguments are not relevant for one-dimensional effects plots. A related problem is that it is not possible to control the graphical options of the layers appearing in the plot. For example:
```{r, message = F, fig.width = 6, fig.height = 4, fig.align='center'}
plot(fitG, select = 3, scale = FALSE, col = "red")
```

makes the whole plot red, because the `col = "red"` argument is passed, via the `...` argument (a.k.a. `ellipsis`), to each plotting function called within `plot.gam` (e.g. `rug`). So it is hard to customize the appearance of the plot. The fact that the graphical elements of the plot are rendered in a fixed order by `plot.gam` is also limiting, for example:
```{r, message = F, fig.width = 8, fig.height = 4, fig.align='center'}
par(mfrow = c(1, 2))
plot(fitG, select = 3, scale = FALSE, residuals = TRUE, ylim = c(-2000, 2000))
plot(fitG, select = 3, scale = FALSE, residuals = TRUE, pch = 1, ylim = c(-2000, 2000))
```

Here the points representing the residuals are always plotted after the fit and confidence intervals so, if you change the size of the points, they will completely hide the fit, as on the plot on the right. We could probably live with the limitations mentioned so far, but the biggest issue is that it is quite hard to add new features to the plots offered by `plot.gam`. In fact, `plot.gam` is a big multi-purpose function, and adding a new graphical features requires looking at its source code, understanding what is going on, and modifing the whole function (e.g. we need to add new arguments, choose at which point within `plot.gam` the new graphical layer should be rendered, etc). So adding new features is a rather involved process. 

The good news is that the limitations just mentioned make of `plot.gam` a very good case study illustrating the usefulness of layer-based framework provided by `ggplot2`. In particular, let's see how the `mgcViz` package solves these issues by adopting the layered system based on `ggplot2`. To use it, we need to convert the output of `gam` to an object of class `gamViz`:
```{r, message = F, warning = FALSE}
library(mgcViz)
fitG_v <- getViz(fitG)

class(fitG)
class(fitG_v)
```
so we see that the output of `getViz` is still of class `gam`. For the purpose of illustration, here we show how to plot an effect step by step, but in practice there are shortcuts (see `?plot.gamViz`). We start by extracting individual effects using the `sm` function:
```{r}
e3 <- sm(fitG_v, 3)
class(e3)
```
So `e3` is a one-dimensional effect built using a thin plate regression splines basis (`tprs`), which we can plot using a class-specific method:
```{r}
pl3 <- plot(e3) # calls plot.mgcv.smooth.1D()
class(pl3)
```
Having stored the plot in `pl3`, an object of class `plotSmooth` which inherits from `gg`, we can then add layers and render
```{r, fig.width = 5, fig.height = 3, fig.align='center'}
pl3 <- pl3 + l_fitLine() + l_ciLine()
pl3 
```

where the last line calls `print.plotSmooth`. In one step:
```{r, fig.show='hide'}
plot(sm(fitG_v, 3)) + l_fitLine() + l_ciLine() # not shown
```
How does this work? The functions starting with `l_` are layers implemented in `mgcViz` and can be seen as wrappers around one or more `ggplot2` layers (e.g. `mgcViz::l_fitLine` is a wrapper around `mgcViz::geom_line`). All `mgcViz` layers start with the `l_` prefix. 
Why is this useful? Because, as we illustrate below:

   - we have now full control over which graphical arguments are passed to each layer;
   - we can add the layers in the order we like;
   - adding new layers does not require modifying the function that created the initial plot (here `plot.mgcv.smooth.1D`)
   - the number of arguments of the initial plotting function is greatly reduced, in fact
```{r}
args( plot.mgcv.smooth.1D )
```

There is a variety of layers available, for example:
```{r plot_sm_lay_one_all, message = F, fig.width = 9, fig.height = 4, fig.align='center'}
pl3a <- plot(sm(fitG_v, 3)) + l_ciPoly(level = 0.99) + l_fitLine(color = "red") + l_rug() 
pl3b <- plot(sm(fitG_v, 3), nsim = 20) + l_simLine()

gridPrint(pl3a, pl3b, ncol = 2)
```

where the second plot shows a set of 20 random curves drawn from (Gaussian approximation to) the posterior distribution of the seasonal effect (but this is not crucial here). As for most smooth effect plots in `mgcViz`, the output of `plot.mgcv.smooth.1D` has class `plotSmooth`, which contains an object of class `c("gg", "ggObj")` is its `$ggObj` slot. Here `gridPrint` simply extracts the `ggplot` objects from `pl3a$ggObj` and `pl3b$ggObj`, and renders them on one page. To get the full list of available `mgcViz` layers do:
```{r listLayers, message = F}
listLayers( plot(sm(fitG_v, 3)) )
```
We can of course use layers and functions provided by `ggplot2`, for example:
```{r ggplot2, message = F, fig.width = 5, fig.height = 3, fig.align='center', warning = FALSE}
pl <- plot(sm(fitG_v, 3))
pl + geom_hex(data = pl$data$res, alpha = 0.7) + l_fitLine() + 
     l_ciLine() + xlim(c(0.25, 0.75))
```

Here we used `ggplot2::geom_hex` to plot the density of the residuals, which required specifying that the residual data can be found in `pl$data$res`, and we zoomed along the $x$-axis on $[0.25, 0.75]$ using `ggplot2::xlim`. 

So this should illustrate some of the advantages of a layered system. To get some insights into how this layered system might be useful to someone trying to build a graphical library for GAM models (but this applies to other models, of course), it's probably instructive to have a look at how `ggplot2` layers can be wrapped up to extend the array of layers made available by `mgcViz`. Here we consider implementing an `mgcViz` layer which plots the residual density using binning, as in the last plot above. Suppose that we want to call such layer `l_binRes`. The first thing we need to do, is to build its general (or generic-like) version:
```{r}
l_binRes <- function(...){
  arg <- list(...)
  o <- structure(list("fun" = "myBinRes",
                      "arg" = arg), 
                 class = "gamLayer")
  return(o)
}
```
This is the template used by all `mgcViz` layers. The layer returns an object of class `gamLayer`, where the `arg` slot contains all arguments that will be passed to the type-specific layer (you'll see what we mean by type-specific in a minute). The `fun` slot indicates the name of the internal function to be used. This is:
```{r}
myBinRes.1D <- function(a){
  
  a$data <- a$data$res
  a$mapping <- aes(x = x, y = y)
  out <- do.call("geom_hex", a)
  return( out )
  
}
```
As you can see, this function returns the output of the `ggplot2::geom_hex` layer. The `.1D` suffix in `myBinRes.1D` is there because it matches the type of the plot we are focussing on:
```{r}
pl$type
```
Having set up the general `l_binRes` function and the internal function `myBinRes.1D`, we can now do:
```{r, fig.align='center'}
pl + l_binRes(alpha = 0.7) + l_fitLine() + l_ciLine()
```

which works! We could say that this is an informal object oriented framework, because we are dispatching the method (`myBinRes.1D`) based on the `type` slot, rather than on a formal class. If we wanted to use the new layer on an plot of a different `type`, we would need to develop a specific `myBinRes` function for that plot type. For instance, if we extract the day-of-week effect and build its plot:
```{r}
pl <- plot(pterm(fitG_v, 1))
```
this would not work:
```{r, fig.show = 'hide'}
pl + l_binRes(alpha = 0.9) + l_fitPoints(col = 2) + l_ciBar(col = 2)
```

To make `binRes` work with plots of type `c("Pterm", "Factor")`, we need to define:
```{r}
myBinRes.PtermFactor <- function(a){
  
  a$data <- a$data$res
  a$mapping <- aes(x = x, y = y, z = y)
  a$fun <- function(x) length(x)
  out <- do.call("stat_summary_2d", a)
  return( out )
  
}
```
where we used the `stat_summary_2d` layers from `ggplot2`, which is more appropriate when the $x$-axis is categorical. Now we get:
```{r, fig.align='center'}
pl + l_binRes(alpha = 0.7) + l_fitPoints(col = 2) + l_ciBar()
```
The point of all this is illustrating that, using the layer-based system provided by `ggplot`, it is easy to introduce new GAM-specific layers for `mgcViz`, without modifying the functions used to build the initial plots (here `plot.mgcv.smooth.1D` and `plot.ptermFactor`). Of course, the new layer introduced here (`l_binRes`) is quite trivial, and we might be better off using the raw `ggplot2` layers directly (e.g. `stat_summary_2d`), but some of the layers implemented by `mgcViz` are quite complicated, hence coding them in `mgcViz` is worthwhile (see, e.g., the code for `mgcViz:::l_fitDens.1D`).




